---
title: "Predictive Liver Disease"
output: html_notebook
---

Libraries and packages

```{r}
library(ggplot2)
library(dplyr)
library(caret)
library(randomForest)
library(car)
library(gridExtra)
library(reshape2)
library(ggcorrplot)
library(corrplot)
library(ggfortify)
```

Download the dataset

```{r}
data <- read.csv("/Users/asus/Desktop/R/Liver_disease_data.csv")
head(data)
summary(data)
```

Visualize the distribution of each variable in your dataset

First, let's do bar plots for categorical variables

```{r}
categorical_vars <- c("Gender", "Smoking", "GeneticRisk", "Diabetes", "Hypertension", "Diagnosis")

# Create individual count plots
plot_list <- lapply(categorical_vars, function(var) {
  ggplot(data, aes_string(x = var)) +
    geom_bar(fill = "coral") +
    theme_minimal() +
    labs(title = paste("Count Plot of", var), x = var, y = "Count")
})

# Arrange plots in a grid
grid.arrange(grobs = plot_list, ncol = 2)
```

Now, histograms for continuous variables 

```{r}
continuous_vars <- c("Age", "BMI", "AlcoholConsumption", "PhysicalActivity", "LiverFunctionTest")

# Create individual histograms
hist_list <- lapply(continuous_vars, function(var) {
  ggplot(data, aes_string(x = var)) +
    geom_histogram(binwidth = 5, fill = "coral", color = "black", alpha = 0.7) +
    theme_minimal() +
    labs(title = paste("Histogram of", var), x = var, y = "Frequency") +
    theme(plot.title = element_text(hjust = 0.5))
})

# Arrange histograms in a grid
grid.arrange(grobs = hist_list, ncol = 2)
```
Matrix of correlation between variables 

```{r}

correlation_matrix <- cor(data[, sapply(data, is.numeric)])

ggcorrplot(
  correlation_matrix,
  type = "full",              
  lab = TRUE,                 # correlation coefficients
  lab_size = 3,               
  tl.cex = 10,                
  colors = c("blue", "white", "red")  
) +
  ggtitle("Correlation Matrix of Variables") +
  theme(plot.title = element_text(hjust = 0.5, size = 14))
```

Plotting each variable against the Diagnosis variable to further investigate how might each variable play a role in the prediction model

```{r}
coral_color <- "#FF7F50"  # Coral color
light_blue <- "#00BFFF"  # Light blue color

# plotting categorical variables against Diagnosis
for (cat_var in categorical_vars) {
  p <- ggplot(data, aes_string(x = cat_var, fill = "factor(Diagnosis)")) +
    geom_bar(position = "dodge") +
    labs(
      title = paste("Distribution of", cat_var, "by Diagnosis"),
      x = cat_var,
      y = "Count",
      fill = "Diagnosis"
    ) +
    scale_fill_manual(values = c(coral_color, light_blue)) + 
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 14))
  print(p)  
}
```

```{r}
# Plotting continuous variables against Diagnosis using boxplots
for (cont_var in continuous_vars) {
  p <- ggplot(data, aes_string(x = "factor(Diagnosis)", y = cont_var, fill = "factor(Diagnosis)")) +
    geom_boxplot() +
    labs(
      title = paste(cont_var, "Distribution by Diagnosis"),
      x = "Diagnosis",
      y = cont_var,
      fill = "Diagnosis"
    ) +
    scale_fill_manual(values = c(coral_color, light_blue)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 14))
  print(p)  # Explicitly print the plot
}
```
Using the combine effect of genetics and lifestyle to explore interactions between predictors.

```{r}
ggplot(data, aes(x = AlcoholConsumption, y = LiverFunctionTest, color = GeneticRisk)) +
  geom_point() +
  facet_wrap(~ Diagnosis) +
  labs(title = "Interaction: AlcoholConsumption vs LiverFunctionTest by GeneticRisk",
       x = "Alcohol Consumption", y = "Liver Function Test")
```
Now let's do some PCA:
See how much of the variance is explained by the first PCs


```{r}
pca_var <- pca_model$sdev^2
pca_var_explained <- pca_var / sum(pca_var)

# scree plot
plot(pca_var_explained, type = "b", pch = 19, col = "blue", 
     xlab = "Principal Component", ylab = "Proportion of Variance Explained",
     main = "Scree Plot")
```

Which variables contribute to each PC


```{r}
# loadings for each variable
print(pca_model$rotation)

# contribution of variables to PC1
loading_PC1 <- pca_model$rotation[,1]
loading_PC1[order(abs(loading_PC1), decreasing = TRUE)]
```













